---
title: Ordinary Least Squares
subtitle: ST303 - Linear Models
filters:
  - code-window
---

```{r}
#| label: setup

set.seed(4321)

# Output -----------------------------------------------------------------------
knitr::opts_chunk$set(fig.align = "center")

# Packages ---------------------------------------------------------------------
library(dplyr)
library(ggplot2)
library(tinytable)

# Theme ------------------------------------------------------------------------
theme_st303 <- function() {
  theme_bw() +
    theme(
      axis.title = element_text(size = rel(0.8), face = "bold"),
      legend.title.position = "top",
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

# Data -------------------------------------------------------------------------
heights <- readRDS(here::here("data/heights.RDS"))
fuel <- readRDS(here::here("data/fuel.RDS"))
```

## Where did we leave off

Earlier today we asked:

::: {.callout-note}
## **What is the "best" line?**
:::

And we decided that the best line is the one that

::: {.callout-tip}
## Minimises the sum of squared residuals
:::

$$
\sum_{i=1}^{n}\epsilon_i^2
$$

<br>

::: {.fragment}
[**Now, we'll be asking how do we actually find that line?**]{.center}
:::

## The Goal

Suppose our regression line is

$$
\hat{y_i} = \beta_0 + \beta_1x_i + \epsilon
$$

Find the values of $\beta_0$ and $\beta_1$ that give the smallest possible residual


:::: {.columns}
::: {.column}
$$
\begin{align*}
S(\beta_0, \beta_1) &= \sum^n_{i = 1} \epsilon_i^2 \\[4pt]
&= \sum^n_{i = 1} (y_i - \hat{y_i})^2 \\[4pt]
&= \boxed{\sum^n_{i = 1} (y_i - \beta_0 - \beta_1x_i)^2}
\end{align*}
$$

:::
::: {.column}
::: {.fragment}
<br>
<br>

**Becomes an optimisation problem**
:::
:::
::::

## The Goal

$$
\boxed{S(\beta_0, \beta_1) = \sum^n_{i = 1} (y_i - \beta_0 - \beta_1x_i)^2}
$$


There are **infinitely** many possible values of $\beta_0$ and $\beta_1$.

Each **combination** gives us a **different line**.

Each line gives us **different residuals**.

Each set of residuals gives us a **different value** of $\beta_0$ and $\beta_1$

<br>

::: {.callout-important icon="false"}
## **We want the values of $\beta_0$ and $\beta_1$ that make $S$ as small as possible.**
:::

---

<br>

```{r}
#| label: fig-interactive-regression
#| fig-cap: "Interactive demonstration of how different slopes affect the candidate line and residuals. Watch how the dashed lines (residuals) change as the line changes."

# Some candiate slope lines
slopes <- seq(0, 1, length.out = 100)

# Predict from all these slopes
anim_data <- purrr::map_dfr(slopes, function(s) {
  b0 <- mean(heights$daughter - s * heights$mother)

  heights |>
    mutate(
      slope = s,
      intercept = b0,
      predicted = b0 + s * mother,
      residual = daughter - predicted
    )
})

# Calculate sum of squared residuals
anim_data <- anim_data |>
  group_by(slope, intercept) |>
  mutate(
    ssr = sum(residual^2)
  ) |>
  ungroup()

# Plot the different lines -----------------------------------------------------
anim_data |>
  ggplot(aes(mother, daughter)) +
  geom_jitter(
    data = heights,
    width = 0,
    height = 0.15
  ) +
  geom_abline(
    aes(slope = slope, intercept = intercept),
    linewidth = 1.2,
    colour = ggokabeito::palette_okabe_ito(3)
  ) +
  geom_segment(
    aes(
      xend = mother,
      y = predicted,
      yend = daughter,
      colour = abs(residual)
    ),
    linetype = "dashed",
    alpha = 0.7
  ) +
  scale_color_gradient(
    low = "grey80",
    high = "steelblue",
    guide = "none"
  ) +
  labs(
    x = "Mother height",
    y = "Daughter height"
  ) +
  theme_st303() +
  gganimate::transition_states(
    slope,
    transition_length = 3,
    state_length = 0.5
  )
```

## Finding $\beta_0$ and $\beta_1$

We want to find the line that gives the **smallest possible sum of squared errors**.

There are two things we can change:
$$
\boxed{\beta_0 = \text{intercept}} \qquad \boxed{\beta_1 = \text{slope}}
$$


::: {.callout-important icon="false"}
## What values of $\beta_0$ and $\beta_1$ give us the minimum error?
:::

::: {.fragment}
Using calculus, we can find the point where changing **either** the
intercept or slope would no longer reduce the error:

$$
\frac{\partial S}{\partial \beta_0}=0
\qquad
\frac{\partial S}{\partial \beta_1}=0
$$
:::

## Finding $\beta_0$ and $\beta_1$ {style="font-size: 75%;"}

### How to derive $\beta_0$

$$
\boxed{S(\beta_0, \beta_1) = \sum^n_{i = 1} (y_i - \beta_0 - \beta_1x_i)^2}
$$

::: {.fragment}
$$
\frac{\partial S}{\partial \beta_0} = -2 \sum^n_{i = 1} (y_i - \beta_0 - \beta_1x_i) \\[8pt]
\sum^n_{i = 1} (y_i - \hat{\beta_0} - \hat{\beta_1}x_i) = 0 \\[8pt]
\sum y_i - n\hat{\beta_0} - \hat{\beta_1}\sum x_i = 0 \\[8pt]
\bar{y} - \hat{\beta_0} - \hat{\beta_1}\bar{x} = 0 \\[8pt]
\boxed{\hat{\beta_0} = \bar{y} - \hat{\beta_1}\bar{x}}
$$

:::
::: {.fragment}
::: {.callout-important}
## The OLS regression line always passes through $(\bar{x},\bar{y})$.
:::
:::

## Finding $\beta_0$ and $\beta_1$ {style="font-size: 75%;"}

### How to derive $\beta_1$

$$
\boxed{S(\beta_0, \beta_1) = \sum^n_{i = 1} (y_i - \beta_0 - \beta_1x_i)^2}
$$

::: {.fragment}
$$
\frac{\partial S}{\partial \beta_0} = -2\sum^n_{i = 1} x_i (y_i - \beta_0 - \beta_1x_i) \\[4pt]
\sum^n_{i = 1} x_i (y_i - \hat{\beta_0} - \hat{\beta_1}x_i) = 0 \\[4pt]
$$

:::

\vspace{-1cm}

::: {.fragment}
$$
\sum^n_{i = 1} x_i (y_i - {\color{blue}\bar{y} - \hat{\beta_1}\bar{x}} - \hat{\beta_1}x_i) = 0 \\[4pt]
\sum^n_{i=1}(x_i - \bar{x})(y_i - \bar{y}) = \hat{\beta_1}\sum^n_{i=1}(x_i - \bar{x})^2 \\[4pt]
\boxed{\hat{\beta_1} = \frac{\sum^n_{i=1}(x_i - \bar{x})(y_i - \bar{y})}{\sum^n_{i=1}(x_i - \bar{x})^2}}
$$

:::


## Some notation

To make this things a little easier to work with, we can define:

$$
S_{xx} = \sum_{i=1}^{n}(x_i-\bar x)^2 \qquad S_{xy} = \sum_{i=1}^{n}(x_i-\bar x)(y_i-\bar y).
$$


::: {.fragment}
Therefore, our derived equations become

$$
\boxed{\hat{\beta_0} = \bar{y} -\hat{\beta_1} \bar{x}} \qquad \boxed{\hat{\beta_1} = \frac{S_{xy}}{S_{xx}}}
$$

:::


::: {.fragment}
::: {.callout-tip}
## At the end of the day, we just need to calculate some means!
:::
:::

## Let's look at an example

Eight office complexes were observed over one week.
We recorded **average temperature** and **fuel consumption**.

```{r}
#| label: tbl-fuel
#| tbl-cap: "Overview of our data"

fuel |>
  rename(Temperature = temp, Fuel = fuel) |>
  tt() |>
  style_tt(i = 0, bold = TRUE) |>
  style_tt(j = c(1, 2), align = "c")
```

---

<br>

```{r}
#| label: fig-fuel-1
#| fig-cap: "Scatter plot showing the relationship between temperature and fuel consumption"
fuel |>
  ggplot(aes(temp, fuel)) +
  geom_point(size = rel(3)) +
  labs(
    x = "Average temperature (°F)",
    y = "Fuel consumption"
  ) +
  theme_st303()
```

::: {.fragment}
::: {.callout-caution icon="false"}
## What do you expect the slope for this data to be?
:::
:::

---

```{.r filename="OLS.R"}
# Deriving beta 0 and beta 1 using a small dataset -----------------------------

# Create a small fuel dataset
fuel <- dplyr::tibble(
  temp = c(28, 28, 32.5, 39, 45.9, 57.8, 58.1, 62.5),
  fuel = c(12.4, 11.7, 12.4, 10.8, 9.4, 9.5, 8, 7.5)
)

# Get our means ----------------------------------------------------------------
x_bar <- mean(fuel$temp)
y_bar <- mean(fuel$fuel)

# Get S_xx ---------------------------------------------------------------------
s_xx <- (fuel$temp - x_bar)^2 |> sum()

# Get S_xy ---------------------------------------------------------------------
s_xy <- ((fuel$temp - x_bar) * (fuel$fuel - y_bar)) |> sum()

# Now we can find beta 1 (S_xy / S_xx) -----------------------------------------
beta_1 <- s_xy / s_xx

# And then we can find beta 0 (y_bar - beta_1 * x_bar) --------------------------
beta_0 <- y_bar - (beta_1 * x_bar)

beta_0 # 15.838
beta_1 # -0.128

# Thus our regression becomes: y = 15.838 - 0.128x
```

## {style="font-size: 60%;"}

**Calculate $\bar{x}$ and $\bar{y}$**

$$
\bar{x} = 43.975 \qquad \bar{y} = 10.2125
$$

**Calculate $S_{xx}$**

$$
S_{xx} = \sum^n_{i = 1} (x_i - \bar{x}) ^2 = 1404.355
$$

**Calculate $S_{xy}$**

$$
S_{xy} = \sum^n_{i = 1}(x_i - \bar{x})(y_i - \bar{y}) = -179.6475
$$

**Calculate $\beta_1$**

$$
\hat{\beta_1} = \frac{S_{xy}}{S_{xx}} = \frac{-179.6475}{1404.355} = -0.128
$$

**Calculate $\beta_0$**

$$
\beta_0 = \bar{y} - \hat{\beta_1}\bar{x} = 10.2125 + 0.128(43.975) = 15.838
$$

$$
\boxed{\hat{y} = 15.838 - 0.128x}
$$

---

<br>

```{r}
#| label: fig-fuel-2
#| fig-cap: "Scatter plot showing the relationship between temperature and fuel consumption"

label <- "y = 15.838 - 0.128x"

fuel |>
  ggplot(aes(temp, fuel)) +
  geom_point(size = rel(3)) +
  geom_smooth(method = "lm", colour = "#56B4E9", se = FALSE) +
  ggtext::geom_richtext(
    data = tibble(x = 52, y = 12),
    aes(x = x, y = y, label = label),
    size = rel(5),
    fontface = "bold"
  ) +
  labs(
    x = "Average temperature (°F)",
    y = "Fuel consumption"
  ) +
  theme_st303()
```

---

<br>

```{r}
#| label: fig-fuel-3
#| fig-cap: "Scatter plot showing the relationship between temperature and fuel consumption"

label <- "10.078 ≈ 15.838 - 0.128(45)"

fuel |>
  ggplot(aes(temp, fuel)) +
  geom_point(size = rel(3)) +
  geom_smooth(method = "lm", colour = "#56B4E9", se = FALSE) +
  geom_point(
    data = tibble(x = 45, y = 10.078),
    aes(x = x, y = y),
    colour = "red",
    size = rel(6)
  ) +
  ggtext::geom_richtext(
    data = tibble(x = 52, y = 12),
    aes(x = x, y = y, label = label),
    size = rel(5),
    fontface = "bold"
  ) +
  labs(
    x = "Average temperature (°F)",
    y = "Fuel consumption"
  ) +
  theme_st303()
```

## Some properties of OLS fits

- The residuals sum to zero $\sum^n_{i=1} \epsilon^2_i = 0$
- The OLS line passed through $(\bar{x}, \bar{y})$
- The observed values and predicted values have the same mean $(\bar{y} = \bar{y_i})$
- Residuals are uncorrelated with $x$ $(\sum x_i \epsilon_i = 0)$

## Practice question

$$
\boxed{\hat{\beta_0} = \bar{y} -\hat{\beta_1} \bar{x}} \qquad \boxed{\hat{\beta_1} = \frac{S_{xy}}{S_{xx}}}
$$


::: {.callout-note icon="false"}
## Find the OLS regression line from the following results

$$
\bar{x} = 10 \qquad \bar{y} = 25 \qquad S_{xx} = 50 \qquad S_{xy} = 20
$$

:::


::: {.fragment}
$$
\beta_1 = \frac{20}{50} = 0.4 \qquad \beta_0 = 25 - 0.4(10) = 21
$$

$$
\boxed{\hat{y} = 21 + 0.4x}
$$


:::


## Some special cases

Consider the following two models

**Intercept only simple linear regression model**

$$
y_i = \beta_0 + \epsilon_i
$$
**No-intercept simple linear regression model**

$$
y_i = \beta_1x + \epsilon_i
$$


::: {.callout-tip icon="false"}
## Find the ordinary least squares estimate for $\beta_0$ and $\beta_1$, respectively
:::


::: {.fragment}
$$
\boxed{S(\beta_0, \beta_1) = \sum^n_{i = 1} (y_i - \hat{y_i})^2}
$$

:::

## Intercept only simple linear regression model

$$
S(\beta_0) = \sum^n_{i = 1} (y_i - \beta_0)^2 \\[8pt]
\frac{dS}{d\beta_0} = -2\sum^n_{i = 1} (y_i - \beta_0) \\[8pt]
\sum^n_{i = 1} (y_i - \hat{\beta_0}) = 0 \\[8pt]
\sum y_i - \sum \hat{\beta_0} = 0 \\[8pt]
\sum y_i - n\hat{\beta_0} = 0 \\[8pt]
\hat{\beta_0} = \frac{\sum y_i}{n} = \bar{y}
$$

## No-intercept simple linear regression model

$$
S(\beta_0) = \sum^n_{i = 1} (y_i - \beta_1x_i)^2 \\[8pt]
\frac{dS}{d\beta_1} = -2\sum^n_{i = 1} x_i(y_i - \beta_1x_i) \\[8pt]
\sum^n_{i = 1} x_i(y_i - \hat{\beta_1}x_i) = 0 \\[8pt]
\sum x_i y_i - \hat{\beta_1} \sum x_i^2 = 0 \\[8pt]
\hat{\beta_1} = \frac{\sum x_i y_i}{\sum x_i^2}
$$

---

<br>

```{r}
#| label: fig-special-cases
#| fig-cap: "Illustration of an intercept only and no-intercept model"
#|
fuel |>
  ggplot(aes(x = temp, y = fuel)) +
  geom_point(size = 0, alpha = 0) +
  geom_abline(
    intercept = 10,
    slope = 0,
    linetype = "dashed",
    colour = ggokabeito::palette_okabe_ito(2, alpha = 0.7)
  ) +
  geom_abline(
    intercept = 0,
    slope = 0.2022612,
    linetype = "dashed",
    colour = ggokabeito::palette_okabe_ito(3, alpha = 0.7)
  ) +
  annotate(
    geom = "text",
    x = 10,
    y = 10.5,
    label = "Intercept only model",
    size = rel(5)
  ) +
  annotate(
    geom = "text",
    x = 32,
    y = 5,
    label = "No-intercept model",
    angle = 35,
    size = rel(5)
  ) +
  ylim(0, 12) +
  xlim(0, 60) +
  labs(x = NULL, y = NULL) +
  theme_st303()
```

## Wrapping up

OLS finds the **best-fitting line**.

$$
\sum_{i=1}^n (y_i-\hat y_i)^2
$$

We can calculate both $\beta_0$ and $\beta_1$ such that

$$
\hat{\beta_0} =\bar{y} - \hat{\beta_1}\bar{x} \qquad \hat{\beta}_1= \frac{S_{xy}}{S_{xx}}
$$

- **When there is only an intercept:** $\hat{\beta_0} = \bar{y}$
- **When there is no intercept:** $\hat{\beta_1} = \frac{\sum x_i y_i}{\sum x_i^2}$
