---
title: ANOVA decomposition
subtitle: ST303 - Linear Models
filters:
  - code-window
---

```{r}
#| label: setup

set.seed(123456789)

# Output -----------------------------------------------------------------------
knitr::opts_chunk$set(fig.align = "center")

# Packages ---------------------------------------------------------------------
library(dplyr)
library(ggplot2)
library(tinytable)

library(rosdata) # Some extra datasets

# Function ---------------------------------------------------------------------
source(here::here("R/make_hibbs_plot.R"))

# Theme ------------------------------------------------------------------------
theme_st303 <- function() {
  theme_bw() +
    theme(
      plot.title = element_text(size = rel(1.1), face = "bold"),
      axis.title = element_text(size = rel(0.8), face = "bold"),
      legend.title.position = "top",
      legend.position = "bottom",
      legend.title = element_text(hjust = 0.5, size = rel(0.8), face = "bold"),
      panel.grid.minor = element_blank()
    )
}

# Data -------------------------------------------------------------------------
data("hibbs") # Election versus economic growth

# Model ------------------------------------------------------------------------
fit <- lm(vote ~ growth, data = hibbs)

# Key values -------------------------------------------------------------------
y_bar <- mean(hibbs$vote)
y_hat <- fitted(fit)
esp <- residuals(fit)

SST <- sum((hibbs$vote - y_bar)^2)
SSR <- sum((y_hat - y_bar)^2)
SSE <- sum(esp^2)

R2 <- SSR / SST

F_obs <- summary(fit)$fstatistic[1]
p_F <- pf(
  F_obs,
  df1 = summary(fit)$fstatistic[2],
  df2 = summary(fit)$fstatistic[3],
  lower.tail = FALSE
)
```


## The Hibbs data

<br>

```{r}
#| label: hibbs-overview

make_hibbs_plot(hibbs = hibbs)
```

## The Hibbs data

<br>

```{r}
#| label: hibbs-plot-1a
fig_1a <- hibbs |>
  ggplot(aes(x = growth, y = vote)) +
  geom_point() +
  scale_x_continuous(labels = scales::label_number(suffix = "%")) +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  labs(
    x = "Average recent growth in personal income",
    y = "Incumbent party's vote share"
  ) +
  theme_st303()

fig_1a
```

## The Hibbs data

<br>

```{r}
#| label: hibbs-plot-1b
#| dev: svg
fig_1b <- hibbs |>
  ggplot(aes(x = growth, y = vote)) +
  geom_point() +
  geom_smooth(
    method = "lm",
    se = FALSE,
    linewidth = 0.8,
    linetype = "dashed",
    colour = ggokabeito::palette_okabe_ito(1)
  ) +
  scale_x_continuous(labels = scales::label_number(suffix = "%")) +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  labs(
    x = "Average recent growth in personal income",
    y = "Incumbent party's vote share"
  ) +
  theme_st303()

fig_1b
```

## How much does our regression model explain?

### The regression model

We have fitted a simple linear regression model

$$
Y_i \sim \beta_0 + \beta_1X_i + \varepsilon_i
$$

- $Y$ = incumbent party vote share
- $X$ = economic growth

Our model gives us a fitted value $\hat{Y_i}$ for each observation

[**But how useful are these fitted values**]{.fragment .fade-up .center}

:::: {.fragment .fade-up}
::: {.callout-important}
## How much of the variation in $Y$ is explained by the regression?
:::
::::

## How much does our regression model explain?

### Ignoring the slope

Suppose we ignored economic growth completely

$$
Y_i \sim \beta_0 + \varepsilon_i
$$

```{r}
#| label: hibbs-plot-1c
fig_1c <- hibbs |>
  ggplot(aes(x = growth, y = vote)) +
  geom_point() +
  geom_abline(
    intercept = mean(hibbs$vote),
    slope = 0,
    linetype = "dashed",
    linewidth = 0.8,
    colour = ggokabeito::palette_okabe_ito(1)
  ) +
  scale_x_continuous(labels = scales::label_number(suffix = "%")) +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  labs(
    x = "Average recent growth in personal income",
    y = "Incumbent party's vote share"
  ) +
  theme_st303()

fig_1c
```

## How much does our regression model explain?

### Including the slope

Naturally, when we include the slope we get a different set of predictions

$$
Y_i \sim \beta_0 + \beta_1X_i + \varepsilon_i
$$


```{r}
#| label: hibbs-plot-1d

fig_1b
```

---

<br>

```{r}
#| label: hibbs-plot-1e

fig_1c <- fig_1c + labs(title = "Intercept only model")
fig_1b <- fig_1b + labs(title = "Standard regression model")

ggpubr::ggarrange(plotlist = c(fig_1c, fig_1b))
```

::: {.callout-important}
## The standard regression has reduced the prediction errors
:::

## Let's break this down

### We have **three quantities** for every observation

For every observation $i$ we have the following


:::: {.columns}

::: {.column width="33%"}

[**The observed value**]{.center}

$$Y_i$$

:::

::: {.column width="33%"}

[**The estimated value**]{.center}

$$\hat{Y_i}$$

:::

::: {.column width="33%"}

[**The residual**]{.center}

$$\varepsilon_i$$

:::
::::

<br>


::: {.fragment .fade-up}

Therefore we can say

$$
\boxed{Y_i = \hat{Y_i} + \varepsilon_i}
$$

:::

## From an observation to a variation {style="font-size: 80%;"}

[**We want to understand the variation in $Y$**]{.center}

$$
Y_i - \bar{Y}
$$


:::: {.fragment .fade-up}

$$
\begin{align*}
Y_i &= \hat{Y_i} + \varepsilon_i \\[4pt]
Y_i - \bar{Y} &= (\hat{Y_i} - \bar{Y}) + \varepsilon_i \\[4pt]
\sum^n_{i = 1}(Y_i - \bar{Y})^2 &= \sum^n_{i = 1}\Big[(\hat{Y_i} - \bar{Y}) + \varepsilon_i\Big]^2 \\[4pt]
\sum(Y_i - \bar{Y})^2 &= \sum(\hat{Y_i} - \bar{Y})^2 + \sum \varepsilon_i^2 + 2 \sum(\hat{Y_i} - \bar{Y}) \varepsilon_i
\end{align*}
$$

$$
\boxed{\sum(Y_i - \bar{Y})^2 = \sum(\hat{Y_i} - \bar{Y})^2 + \sum \varepsilon_i^2}
$$


::: {.callout-tip}
## We call this ANOVA decomposition
:::
::::

## From an observation to a variation

We have decomposed the total variation in $Y$ into two parts:

:::: {.columns}
::: {.column}

[**Explained**]{.center}

$$
\sum (\hat{Y}_i - \bar{Y})^2
$$

:::

::: {.column}

[**Unexplained**]{.center}

$$
\sum \varepsilon_i^2
$$

:::
::::

- **Explained**: Variation accounted for by the regression
- **Unexplained**: Variation left over in the residuals

::: {.callout-tip}
## Total variation = Explained variation + Unexplained variation
:::

## Sums of squares {style="font-size: 95%;"}

### These three formulas have specific names

::: {.callout-tip}
## Total variation = Explained variation + Unexplained variation
:::

<br>

| Quantity          | Definition                                   | Meaning               |
| ----------------- | ---------------------------------------------| --------------------- |
| **Total SS**      | $\displaystyle SS_T=\sum(Y_i-\bar Y)^2$      | Total variation       |
| **Regression SS** | $\displaystyle SS_R=\sum(\hat Y_i-\bar Y)^2$ | Explained variation   |
| **Error SS**      | $\displaystyle SS_E=\sum e_i^2$              | Unexplained variation |



::: {.fragment .fade-up}

So another way of writing this

$$
SS_T = SS_R + SS_E
$$

[This is the ANOVA decomposition]{.center .alert}
:::

## The ANOVA table

The decomposition can be summarised as an ANOVA table

<br>

```{r}
#| label: anova-table

anova_data <- data.frame(
  Source = c("Regression", "Error", "**Total**"),
  df = c("$1$", "$n-2$", "$n-1$"),
  SS = c("$SS_R$", "$SS_E$", "$SS_T$"),
  MS = c("$MS_R$", "$MS_E$", ""),
  F = c("$MS_R/MS_E$", "", "")
)

# Output the table
anova_data |> tt(width = 0.75) |> format_tt(markdown = TRUE)
```

where

$$
MS_R = \frac{SS_R}{1} \qquad MS_E = \frac{SS_E}{n - 2}
$$

::: {.notes}
The ANOVA table contains four pieces of information

- degrees of freedom
- sums of squares
- mean squares
- the F-statistic
:::

## The ANOVA table

```{r}
#| label: anova-table-code-1
#| echo: true
#| collapse: true
anova(fit) # ANOVA table for our fitted model
```


::::: {.fragment .fade-up}
:::: {.columns}
::: {.column width="70%"}

- **Explained variation**: 273.63
- **Residual variation**: 198.27
- **Total variation**: 273.63 + 198.27 = 472.91

:::

::: {.column width="30%"}

**How much <br> variation <br> is explained?**

:::

::::
:::::

## How much variation is explained?

$$
SS_T = SS_R + SS_E
$$

[**What proportion of the total variation is explained by the regression**]{style="font-size: 90%;" .center}


::: {.fragment .fade-up}

$$
R^2 = \frac{SS_R}{SS_T} \qquad \text{OR} \qquad R^2 = 1 - \frac{SS_E}{SS_T}
$$



- $R^2 = 0$: 0% of the variation is explained by the regression
- $R^2 = 0.5$: 50% of the variation is explained by the regression
- $R^2 = 1$: 100% of the variation is explained by the regression

:::

::: {.notes}

- $SS_R$ tells us how much variation has been explained.
- $SS_T$ tells us how much variation there was to begin with.

:::

## How much variation is explained?

```{r}
#| label: anova-table-code-2
#| echo: true
#| collapse: true
anova(fit) # ANOVA table for our fitted model
```

$$
R^2 = \frac{SS_R}{SS_T} = \frac{273.63}{472.91} = 0.58
$$

## Hypothesis test {style="font-size: 80%;"}

::: {.callout-caution appearance="minimal"}
## Now we know much variation does the regression explain?
:::

::: {.callout-important appearance="minimal"}
## Is there evidence that the regression explains more variation than we would expect if there were no linear relationship?
:::

$$
H_0: \beta_1 = 0 \qquad H_1: \beta_1 \ne 0
$$


::: {.fragment .fade-up}

- If $\beta_1 = 0$ then there is no linear contribution from $X$
- The fitted regression should therefore provide little improvement over the
  <br> intercept-only model.
- We believe the population model is essentially

$$
Y_i \sim \beta_0 + \varepsilon_i
$$

:::


::: {.fragment .fade-up}
Hence, if $H_0$ was true, we would expect $SS_R$ to be relatively small
compared to $SS_E$
:::

## Hypothesis test {style="font-size: 80%;"}

### F statistic

We compare the two mean squares

$$
F = \frac{MS_R}{MS_E} \qquad \text{OR} \qquad F = \frac{SS_R/1}{SS_E/(n-2)}
$$

If $H_0$ was true, we would expect $F \approx 1$


:::: {.fragment .fade-up}

If the regression explains substantially more variation than expected under $H_0$

$$
MS_R \gg MS_E \qquad \text{Then} \qquad F \gg 1
$$

Hence, large F values provide evidence against $H_0$


::: {.callout-important}
## But how do we determine a large F value?
:::
::::

## Hypothesis test

### F distribution

The F statistic has it's own corresponding distribution
[the F distribution]{.alert}

```{r}
#| label: f-distribution-1
fig_2 <- tibble(f = rf(n = 1000, df1 = 1, df2 = 14)) |>
  ggplot(aes(x = f)) +
  geom_density(colour = "#E69F00", linewidth = 0.8) +
  labs(x = NULL, y = "Density") +
  theme_st303()

fig_2
```

## Hypothesis test

### F distribution

The F statistic has it's own corresponding distribution
[the F distribution]{.alert}

```{r}
#| label: f-distribution-2
fig_2 + geom_vline(xintercept = 19.321, linetype = "dashed", colour = "#999999")
```

## Wrapping up {style="font-size: 80%;"}

:::: {.columns}

::: {.column width="33%"}

[**Decomposition**]{.center}

$$
\boxed{SS_T + SS_R + SS_E}
$$

:::

::: {.column width="33%"}

[**Variation explained**]{.center}

$$
\boxed{R^2 = \frac{SS_R}{SS_T}}
$$

:::

::: {.column width="33%"}

[**F-statistic**]{.center}

$$
\boxed{F = \frac{MS_R}{MS_E}}
$$

:::
::::

<br>

ANOVA decomposition gives us a way to think about regression as a
decomposition of variation

::: {.callout-tip}
## Total variation = Explained variation + Unexplained variation
:::

<br>

:::: {.columns}

::: {.column width="33%"}
[Simple linear regression]{.center}
:::

::: {.column width="33%"}
[Multiple linear regression]{.center}
:::

::: {.column width="33%"}
[Comparing models]{.center}
:::
::::
