---
title: The Linear Model
subtitle: ST303 - Linear Models
filters:
  - code-window
---
```{r}
#| label: setup

set.seed(4321)

# Options ----------------------------------------------------------------------
knitr::opts_chunk$set(fig.align = "center")

# Packages ---------------------------------------------------------------------
library(dplyr)
library(ggplot2)
library(tinytable)

# Theme ------------------------------------------------------------------------
theme_st303 <- function() {
  theme_bw() +
    theme(
      axis.title = element_text(size = rel(0.8), face = "bold"),
      legend.title.position = "top",
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

# Data -------------------------------------------------------------------------
heights <- readRDS(here::here("data/heights.RDS"))
bacteria <- readRDS(here::here("data/bacteria.RDS"))
```

<br>
<br>
<br> <!--- EXTREMELY HACKY -->
<br>
<br>

[**How can we use data to understand and predict numerical outcomes?**]{.center style="font-size: 90%;"}

## Let's start with a question

[**Can we predict the height of someone's daughter?**]{.center}

::: {.fragment .fade-up}
::: {.callout-note icon="false"}
## Suppose we know the height of someone's mother.
:::

::: {.callout-note icon="false"}
## Can we use that information to predict her height?
:::
:::

<br>

::: {.fragment .fade-up}
::: {.callout-tip icon="false"}
## What would you expect?
:::
:::

::: {.fragment .fade-up}
::: {.callout-tip icon="false"}
## Would you expect a taller mother to have a taller daughter?
:::
:::

::: {.fragment .fade-up}
::: {.callout-tip icon="false"}
## Would you expect the relationship to be perfect?
:::
:::

::: {.fragment .fade-up}
::: {.callout-tip icon="false"}
## How accurate do you think our predictions could be?
:::
:::

::: {.notes}
- Try and answer the following questions
:::

## Let's start with a question

[**Can we predict the height of someone's daughter?**]{.center}

::: {.callout-note icon="false"}
## Suppose we know the height of someone's mother.
:::

::: {.callout-note icon="false"}
## Can we use that information to predict her height?
:::

<br>

::: {.callout-caution icon="false" title="Let's do a quick exercise"}
Let's say the height of a mother is 64.37 inches.

[**What would you predict the height of the daughter to be** ]{.center}
:::

---

<br>

```{r}
#| label: fig-heights-1
#| fig-cap: "Height of mother versus height of daughter"

heights |>
  ggplot(aes(x = mother, y = daughter)) +
  geom_jitter() +
  scale_x_continuous(labels = scales::label_number(suffix = "in")) +
  scale_y_continuous(labels = scales::label_number(suffix = "in")) +
  labs(x = "Mother height", y = "Daughter height") +
  theme_st303()
```

::: {.notes}
- Is there a relationship?
- What direction is it?
- How strong does it appear?
- Does it look linear?
- Are there any unusual observations?
:::

## Predict the height of someone's daughter

```{r}
#| label: heights

heights |>
  ggplot(aes(x = mother, y = daughter)) +
  geom_jitter() +
  scale_x_continuous(labels = scales::label_number(suffix = "in")) +
  scale_y_continuous(labels = scales::label_number(suffix = "in")) +
  labs(x = "Mother height", y = "Daughter height") +
  theme_st303()
```

We have two variables

$$
X = \text{Mother's height} \\[8pt]
Y = \text{Daughter's height}
$$

::: {.callout-note icon="false"}
## Could we summarise this data using a single line?
:::

---

<br>

```{r}
#| label: fig-best-line-example
#| fig-cap: "Height of mother versus height of daughter (with prediction lines)"

# Bunch of different prediction lines
heights |>
  ggplot(aes(x = mother, y = daughter)) +
  geom_jitter() +
  geom_line(
    aes(x = mother, y = mother + 5),
    colour = ggokabeito::palette_okabe_ito(2)
  ) +
  geom_line(
    aes(x = mother, y = mother - 1.5),
    colour = ggokabeito::palette_okabe_ito(6)
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    colour = ggokabeito::palette_okabe_ito(3)
  ) +
  scale_x_continuous(labels = scales::label_number(suffix = "in")) +
  scale_y_continuous(labels = scales::label_number(suffix = "in")) +
  labs(x = "Mother height", y = "Daughter height") +
  theme_st303()
```

::: {.notes}
- Here, I've drawn three lines through the data
- Which provides the most accurate summation of the data
:::

## But what is the best line?

<br>

```{r}
#| label: residual-example
#|
heights |>
  mutate(
    to_colour = case_when(
      mother == 60 & daughter == 61 ~ TRUE,
      TRUE ~ FALSE
    )
  ) |>
  ggplot(aes(x = mother, y = daughter, colour = to_colour)) +
  geom_jitter(aes(size = to_colour)) +
  geom_line(
    aes(x = mother, y = mother + 5),
    colour = ggokabeito::palette_okabe_ito(2)
  ) +
  geom_line(
    aes(x = mother, y = mother - 1.5),
    colour = ggokabeito::palette_okabe_ito(6)
  ) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    colour = ggokabeito::palette_okabe_ito(3)
  ) +
  scale_x_continuous(labels = scales::label_number(suffix = "in")) +
  scale_y_continuous(labels = scales::label_number(suffix = "in")) +
  scale_colour_manual(
    values = c(
      "TRUE" = ggokabeito::palette_okabe_ito(7),
      "FALSE" = ggokabeito::palette_okabe_ito(8, alpha = 0.5)
    )
  ) +
  scale_size_manual(
    values = c(
      "TRUE" = rel(3.5),
      "FALSE" = rel(0.8)
    )
  ) +
  labs(x = "Mother height", y = "Daughter height") +
  guides(colour = "none", size = "none") +
  theme_st303()
```

::: {.notes}
- But how do we know this
- Obviously, we could just eye ball it
- But we need a way to quantitively prove the green line is the best line
- This is where **residuals** come in
:::

## But what is the best line?

```{r}
tribble(
  ~""             , ~Mother , ~Daughter ,
  "Original data" ,      60 , 61        ,
  "Blue line"     ,      60 , 65        ,
  "Orange line"   ,      60 , 58.5      ,
  "Green line"    ,      60 , 60.74
) |>
  tt() |>
  style_tt(i = 0, bold = TRUE) |>
  style_tt(j = c(2, 3), align = "c")
```

::: {.fragment}
::: {.callout-note icon="false"}
## Blue over predicted by 4 inches
:::

::: {.callout-warning icon="false"}
## Orange under predicted by 2.5 inches
:::

::: {.callout-tip icon="false"}
## Green under predicted by 0.26 inches
:::
:::

::: {.fragment .fade-up}
::: {.callout-important icon="false"}
## We call these differences "residuals"

$$
\epsilon_i = y_i - \hat{y}_i
$$
:::
:::

## But what is the best line?

### Residuals

Every possible line gives us a different set of residuals.

**Residuals hold the key to determining which line is the best.**

:::: {.fragment .fade-up}
We want a line that **minimises the sum of squared residuals**.

$$
\text{minimise} \sum^n_{i = 1}(y_i - \hat{y}_i)^2
$$

::: {.callout-note icon="false"}
## This is the idea behind ordinary least squares.
:::
::::

---

```{r}
#| label: fig-interactive-regression
#| fig-cap: "Interactive demonstration of how different slopes affect the candidate line and residuals. Watch how the dashed lines (residuals) change as the line changes."

# Some candiate slope lines ----------------------------------------------------
slopes <- seq(0, 1, length.out = 100)

# Predict from all these slopes ------------------------------------------------
anim_data <- purrr::map_dfr(slopes, function(s) {
  b0 <- mean(heights$daughter - s * heights$mother)

  heights |>
    mutate(
      slope = s,
      intercept = b0,
      predicted = b0 + s * mother,
      residual = daughter - predicted
    )
})

# Calculate sum of squared residuals -------------------------------------------
anim_data <- anim_data |>
  group_by(slope, intercept) |>
  mutate(ssr = sum(residual^2)) |>
  ungroup()

# Plot the different lines -----------------------------------------------------
anim_data |>
  ggplot(aes(mother, daughter)) +
  geom_jitter(
    data = heights,
    width = 0,
    height = 0.15
  ) +
  geom_abline(
    aes(
      slope = slope,
      intercept = intercept
    ),
    linewidth = 1,
    colour = ggokabeito::palette_okabe_ito(3)
  ) +
  geom_segment(
    aes(
      xend = mother,
      y = predicted,
      yend = daughter
    ),
    colour = "grey70",
    linetype = "dashed"
  ) +
  labs(
    x = "Mother height",
    y = "Daughter height"
  ) +
  theme_st303() +
  gganimate::transition_states(
    slope,
    transition_length = 2,
    state_length = 1
  )
```

## The regression equation

$$
\hat{y_i} = \beta_0 + \beta_1x_i + \epsilon_i
$$


::: {.fragment}
- $\hat{Y}$ is our **predicted value**
- $X$ is our **observed value**
- $\beta_0$ is our **intercept**
- $\beta_1$ is our **slope**
- $\epsilon$ is our **error term** (we'll come back to this)
:::

:::: {.fragment .fade-up}
::: {.callout-note title="You might recognise"}
This is basically the same as the $y = mx + c$ equation you may have learned
growing up
:::
::::

::: {.notes}
- Same as $y = mx + c$
- We've just complicated things by adding in Greek letters and an error term
:::

## The regression equation

### The slope

Let's say we have the following regression equation

$$
\hat{y} = 30 + \textbf{0.5}x
$$

:::: {.fragment}
::: {.callout-note icon="false"}
## What does 0.5 mean?
:::
::::

:::: {.fragment}
::: {.callout-tip icon="false"}
## For every 1-unit increase in X, predicted values of Y increases by 0.5 units.
:::
::::

<br>

:::: {.fragment}
::: {.callout-caution icon="false"}
## A higher slope value means that the predicted value of Y increases faster
:::
::::

---

<br>

```{r}
#| label: slope-example
#|
tibble(
  x = seq(1:100),
  y_1 = 30 + (0.5 * x),
  y_2 = 30 + (1 * x),
  y_3 = 30 + (3 * x)
) |>
  tidyr::pivot_longer(
    cols = !x,
    names_to = "slope",
    values_to = "y"
  ) |>
  mutate(
    slope = case_when(
      slope == "y_1" ~ "Slope = 0.5",
      slope == "y_2" ~ "Slope = 1",
      slope == "y_3" ~ "Slope = 2",
    )
  ) |>
  ggplot(aes(x = x, y = y, colour = slope)) +
  geom_line(linewidth = rel(1)) +
  scale_y_continuous(breaks = seq(30, 330, by = 70)) +
  ggokabeito::scale_colour_okabe_ito() +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme_st303()
```

## The regression equation {style="font-size: 90%;"}

### The intercept

Let's say we have the following regression equation

$$
\hat{y} = \textbf{30} + 0.5x
$$

::: {.callout-note icon="false"}
## But what about the value of 30
:::

::: {.fragment}
::: {.callout-tip icon="false"}
## This is the predicted of Y when X is equal to 0
:::
:::

<br>

:::: {.fragment}
::: {.callout-important icon="false" title="Helpful way to understand"}
A helpful way that I learned to understand this was to think about when you
start a job.

Everyone will have **0 years of experience** when the first start and
will earn the **same salary** (this is the **intercept**)

As their **years of experience progress** their salary will **increase by a set
amount** (this is the **slope**)
:::
::::

::: {.notes}
- However, at times the intercept will have little meaning for us.
- 0 might not be a possible value in the data (you can 0 inches tall)
- Mathematically it needs to be there.
- In reality, it does not always need to be interpreted
:::

---

<br>

```{r}
#| label: intercept-example
#|
tibble(
  x = seq(1:100),
  y_1 = 30 + (0.5 * x),
  y_2 = 30 + (1 * x),
  y_3 = 30 + (3 * x)
) |>
  tidyr::pivot_longer(
    cols = !x,
    names_to = "slope",
    values_to = "y"
  ) |>
  mutate(
    slope = case_when(
      slope == "y_1" ~ "Slope = 0.5",
      slope == "y_2" ~ "Slope = 1",
      slope == "y_3" ~ "Slope = 2",
    )
  ) |>
  ggplot(aes(x = x, y = y, colour = slope)) +
  geom_line(linewidth = rel(1)) +
  geom_point(
    data = tibble(x = 0, y = 30),
    aes(x = x, y = y),
    colour = "black",
    size = rel(5)
  ) +
  scale_y_continuous(breaks = seq(30, 330, by = 70)) +
  labs(x = NULL, y = NULL, colour = NULL) +
  theme_st303()
```

## Back to the mother and daughter example

### How do we find $\beta_0$ and $\beta_1$

```{.r filename="linear-model.R"}
fit <- lm(daughter ~ mother, data = heights)

coef(fit)
```

```{r}
#| label: linear-model-example
#|
fit <- lm(daughter ~ mother, data = heights)

coef(fit)
```

:::: {.fragment .fade-up}
::: {.callout-tip icon="false" title="Model results"}

- $\beta_0$ = 17.61
- $\beta_1$ = 0.74

:::
::::

---

<br>
<br>
<br>
<br>
<br>

[**But where do these values come from?**]{.center}
