---
title: Simple linear regression models
subtitle: ST303 - Linear Models
filters:
  - code-window
---

```{r}
#| label: setup

set.seed(4321)

# Output -----------------------------------------------------------------------
knitr::opts_chunk$set(fig.align = "center")

# Packages ---------------------------------------------------------------------
library(dplyr)
library(ggplot2)
library(tinytable)

# Theme ------------------------------------------------------------------------
theme_st303 <- function() {
  theme_bw() +
    theme(
      axis.title = element_text(size = rel(0.8), face = "bold"),
      legend.title.position = "top",
      legend.position = "bottom",
      legend.title = element_text(hjust = 0.5, size = rel(0.8), face = "bold"),
      panel.grid.minor = element_blank()
    )
}

# Data -------------------------------------------------------------------------
fuel <- readRDS(here::here("data/fuel.RDS"))

# Fit linear model -------------------------------------------------------------
fit <- lm(fuel ~ temp, data = fuel)

n <- nrow(fuel)
beta_0 <- coef(fit)[1]
beta_1 <- coef(fit)[2]
sigma_sq <- sum(residuals(fit)^2) / (n - 2)
```

## Recap {style="font-size: 95%;"}

In the last two lectures we have:

- Used a line to describe the relationship between two numerical variables
- Defined **residuals** as the difference between observed and predicted values
- Chosen the line that **minimises the sum of squared residuals**
- Derived the OLS estimates

$$
\boxed{
    \hat{\beta}_0 = \bar{y} - \hat{\beta}_1 \bar{x} \qquad
    \hat{\beta}_1 = \frac{S_{xy}}{S_{xx}}
  }
$$

:::: {.fragment .fade-up}
::: {.callout-important}
## Check in: Was there anything that was not clear or confusing from last week?
:::
::::

---

## Today

Today we will take a look at

- The simple linear regression model in more detail
- Some assumptions about our error term $(\varepsilon)$
- The sampling distributions of our parameters $(\beta_0, \beta_1, \sigma^2)$

---

[**We will continue using our fuel-consumption data.**]{.center}

```{r}
#| label: fig-fuel-1
#| fig-cap: "Scatter plot showing the relationship between temperature and fuel consumption"
fuel |>
  ggplot(aes(temp, fuel)) +
  geom_point(size = rel(3)) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    colour = ggokabeito::palette_okabe_ito(2)
  ) +
  labs(
    x = "Average temperature (°F)",
    y = "Fuel consumption"
  ) +
  theme_st303()
```


:::: {.fragment .fade-up}
::: {.callout-note}
## Question: Does temperature completely determine fuel consumption?
:::
::::

:::{.notes}
- "All models are wrong, but some are useful" - George Box (1976)
:::

## A line doesn't explain everything

[**Why don't all office complexes use exactly the amount of fuel predicted by the line?**]{style="font-size: 80%;" .center}

::: {.incremental}
- Temperature is **not the only factor** affecting fuel use
- **Buildings differ** from one another
- Measurements are **not perfect**
- There is simply **natural/random variation**
:::

:::: {.fragment .fade-up}
::: {.callout-note}
## The line describes the expected value of $Y$ for a given $X$.
:::

::: {.callout-note}
## The remaining variation is not explained by the line.
:::
::::

## Same temperature, different outcomes {style="font-size: 90%;"}

Suppose **two office complexes** both have a temperature of $40^\circ \text{F}$.

$$
\mathbb{E}[Y \; \vert \; X = 40] = 10.72
$$

::: {.fragment .fade-up}

- However, one complex might use **11.3 units of fuel**.
- Another complex might use **10.1 units of fuel**.
- Because temperature **is not the only thing** affecting fuel use.

:::

:::: {.fragment .fade-up}

:::{.callout-caution}
## Type of building
:::

:::{.callout-caution}
## Insulation
:::

:::{.callout-caution}
## Cost of fuel
:::

::::

## The basic idea

For any given value $x$, we can think of fuel consumption as having two parts:

$$
y_i \sim \text{Systematic part } + \text{ Random part}
$$

:::{.fragment .fade-up}
In the case of our simple linear regression model

$$
\underbrace{\beta_0+\beta_1x_i}_{\text{systematic part}}
\qquad + \qquad
\underbrace{\epsilon_i}_{\text{random part}}
$$

- $\beta_0 + \beta_1x_i$ tells us **what we expect**
- $\epsilon_i$ tells us **how much this observation differs from what we expect**

:::

---

<br>

:::: {.callout-important icon="false"}
## Population model versus fitted model

The line we estimate from our data is only an
**estimate of an underlying population relationship**.

$$
y_i = \beta_0 + \beta_1x_i + \epsilon_i \quad \epsilon_i \sim \mathcal{N}(0, \sigma^2)
$$

::: {.fragment .fade-up}

- $\beta_0$ = Expected value of $y$ when $x$ = 0
- $\beta_1$ = Expected change in $y$ for a unit increase in $x$
- $\sigma^2$ = The variance of the errors

[**These are called population parameters**]{.center}

[**Unfortunately, we usually do not know these values**]{.center}
:::

::: {.fragment .fade-up}

What we do instead is get an **estimate of these parameters**

$$
\hat{y_i} = \hat{\beta_0} + \hat{\beta_1}x_i
$$

:::

::::

---

```{r}
fuel |>
  ggplot(aes(temp, fuel)) +
  geom_point(size = rel(3)) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    colour = ggokabeito::palette_okabe_ito(2)
  ) +
  labs(
    x = "Average temperature (°F)",
    y = "Fuel consumption"
  ) +
  theme_st303()
```


:::: {.callout-note title="Interpretation"}

$$
y_i \sim 15.84 - 0.128x_i
$$

For each one-degree increase in temperature, predicted fuel consumption
decreases by approximately $0.128$ units, on average.

::: {.fragment .fade-up}
[**In this sample**]{.center}
:::

::::

## Errors versus residuals

The **population error** can be represented as

$$
\varepsilon_i = y_i - (\beta_0 + \beta_1x_i)
$$

- But we do not know what $\beta_0$ or $\beta_1$ are
- So we can never observe $\varepsilon_i$
- Instead we calculate a **residual**

$$
e_i = y_i - \hat{y_i}
$$


## Errors versus residuals


:::: {.columns}

::: {.column}

[**Error**]{.center}

$$
\varepsilon_i = y_i - (\beta_0 + \beta_1x_i)
$$

- Population quantity
- Unobservable
- From unknown parameters

:::

::: {.column}

[**Residual**]{.center}

$$
e_i = y_i - \hat{y_i}
$$

- Sample quantity
- Observable
- From fitted parameters

:::

::::


:::: {.fragment .fade-up}

::: {.callout-important}
## Errors belong to the population model
:::


::: {.callout-important}
## Residuals come from our fitted model.
:::

::::

## Why does this matter

To actual use our model, we make several **assumptions** about the errors.

:::{.incremental}
- Centered around 0
- Roughly equally variable
- Independent
- Normally distributed
:::

## Some assumptions about the errors

### Errors have mean zero

$$
\mathbb{E}[\epsilon_i]=0
$$

This means that, on average, the model does not systematically over-predict or
under-predict.

$$
\epsilon_i < 0 \longrightarrow \text{Too low} \qquad \epsilon_i > 0 \longrightarrow \text{Too high}
$$


[**Over many observations, these balance out**]{.center}

## Some assumptions about the errors

### Constant variance

$$
\text{Var}(\epsilon_i)=\sigma^2
$$

The amount of variation around the regression line is assumed
to be roughly the same across values of $X$.

```{r}
#| label: constant-variance

x <- rep(c(20, 40, 60), each = 1000)

# Constant variance
constant <- tibble(x = x, error = rnorm(length(x), 0, 1))

# Plotting
constant |>
  ggplot(aes(x = error, fill = factor(x))) +
  geom_density(alpha = 0.5) +
  ggokabeito::scale_fill_okabe_ito() +
  labs(
    x = "Error",
    y = "Density",
    fill = "Temperature"
  ) +
  facet_wrap(~x, nrow = 1) +
  guides(fill = "none") +
  theme_st303()
```

## Some assumptions about the errors

### Constant variance

<br>

```{r}
#| label: different-variance

x <- rep(c(20, 40, 60), each = 1000)

# Different levels of variance
variance <- tibble(
  x = x,
  constant = rnorm(length(x), 0, 1),
  increasing = case_when(
    x == 20 ~ rnorm(length(x), 0, 1),
    x == 40 ~ rnorm(length(x), 0, 1.5),
    x == 60 ~ rnorm(length(x), 0, 2)
  ),
  decreasing = case_when(
    x == 20 ~ rnorm(length(x), 0, 1),
    x == 40 ~ rnorm(length(x), 0, 0.5),
    x == 60 ~ rnorm(length(x), 0, 0.2)
  )
)

# Cleaning up the data
variance <- variance |>
  tidyr::pivot_longer(cols = !x, names_to = "variance", values_to = "error") |>
  mutate(
    x = factor(x),
    variance = factor(
      variance,
      levels = c("constant", "increasing", "decreasing")
    )
  )

# Plotting
variance |>
  ggplot(aes(x = error, fill = x)) +
  geom_density(alpha = 0.5) +
  ggokabeito::scale_fill_okabe_ito() +
  labs(
    x = "Error",
    y = "Density",
    fill = "Temperature"
  ) +
  facet_grid(
    variance ~ x,
    scales = "free",
    labeller = labeller(
      variance = function(x) stringr::str_to_sentence(x)
    )
  ) +
  guides(fill = "none") +
  theme_st303()
```

## Some assumptions about the errors

### Independence

The error for one observation **should not tell us anything** about the error
for another observation.

$$
\text{Cov}(\epsilon_i,\epsilon_j)=0 \qquad i\neq j
$$


::: {.callout-note icon="false"}
## If Monday's fuel use was unusually high, it should not anything about Tuesday's?
:::

## Some assumptions about the errors

### Normally distributed errors

$$
\epsilon_i \sim N(0,\sigma^2)
$$

Errors are normally distributed with a mean of 0 and a variance of $\sigma^2$

```{r}
#| label: normal-errors
tibble(x = rnorm(n = 10000, mean = 0, sd = 1)) |>
  ggplot(aes(x = x)) +
  geom_density(fill = ggokabeito::palette_okabe_ito(1, alpha = 0.75)) +
  theme_void()
```


:::: {.fragment .fade-up}
::: {.callout-important}
## Normality matters mainly for the exact small-sample inference that comes later.
:::
::::

---

## Putting it all together

Our simple linear regression model is

$$
y_i \sim \beta_0 + \beta_1 x_i + \varepsilon_i
$$

with errors that are

$$
\begin{align*}
  \mathbb{E}[\varepsilon_i] &= 0 \\[8pt]
  \text{Var}(\varepsilon_i) &= \sigma^2 \\[8pt]
  \text{Cov}(\varepsilon_i, \varepsilon_j) &= 0 \\[8pt]
  \varepsilon_i &\sim N(0, \sigma^2)
\end{align*}
$$


:::: {.fragment .fade-up}
::: {.callout-important}
## Errors are centered + equally spread + independent + approximately normally
:::
::::

## Population versus sample model {style="font-size: 90%;"}

The population model is

$$
Y_i = \beta_0 + \beta_1x_i + \epsilon_i
$$

But we don't know

$$
\beta_0,\quad \beta_1,\quad \sigma^2
$$

[**So we use our sample data to estimate them.**]{.center}

Which gives us

$$
\hat\beta_0,\qquad \hat\beta_1,\qquad \hat\sigma^2
$$


:::: {.fragment .fade-up}
::: {.callout-important icon="false"}
## Different samples will give different estimates.
:::
::::

## Imagine repeating the study

Suppose we could repeatedly collect new samples from the same population.

Each sample would give us:

- A different dataset
- A different fitted line
- A different $\hat\beta_0$
- A different $\hat\beta_1$

::: {.callout-important icon="false"}
## The estimates themselves have distributions.
:::

---

## Let's look at the original data

```{.r filename="coeffs.R"}
fit <- lm(fuel ~ temp, data = fuel)

n <- nrow(fuel)
beta_0 <- coef(fit)[1]
beta_1 <- coef(fit)[2]
sigma_sq <- sum(residuals(fit)^2) / (n - 2)
```

$$
N = 8 \qquad \hat{\beta_0} = 15.84 \qquad \hat{\beta_1} = -0.128 \qquad \hat{\sigma^2} = 0.428
$$

## Now let's simulate a dataset {style="font-size: 80%;"}

```{.r filename="simulation.r"}
set.seed(54321) # For reproducibility

x <- fuel$temp
esp <- rnorm(n = n, mean = 0, sd = sqrt(sigma_sq))

# Now let's calculate y
y <- beta_0 + (beta_1 * x) + esp
```

```{r}
#| label: simulation-example

set.seed(54321) # For reproducibility

x <- fuel$temp
esp <- rnorm(n = n, mean = 0, sd = sqrt(sigma_sq))

# Now let's calculate y
y <- beta_0 + (beta_1 * x) + esp
```

:::: {.columns}
::: {.column}

```{r}
#| label: tbl-og-data
#| tbl-cap: "Original fuel data"
fuel |>
  rename(Temperature = temp, Fuel = fuel) |>
  tt() |>
  style_tt(i = 0, bold = TRUE) |>
  style_tt(j = c(1, 2), align = "c")
```

:::
::: {.column}

```{r}
#| label: tbl-sim-data
#| tbl-cap: "Simulated fuel data"
#|
sim_data <- tibble(
  temp = x,
  fuel = y |> round(1)
)

sim_data |>
  rename(Temperature = temp, Fuel = fuel) |>
  tt() |>
  style_tt(i = 0, bold = TRUE) |>
  style_tt(j = c(1, 2), align = "c")
```

:::
::::

## Let's compare our coefficients

**Original model**

$$
\hat{\beta_0} = 15.84 \qquad \hat{\beta_1} = -0.128 \qquad \hat{\sigma^2} = 0.428
$$

```{.r filename="simulated-regression.R"}
# Fit and extract coefficients
sim_fit <- lm(fuel ~ temp, data = sim_data)

beta_0_sim <- coef(sim_fit)[1]
beta_1_sim <- coef(sim_fit)[2]
sigma_sq_sim <- sum(residuals(sim_fit)^2) / (8 - 2)
```

```{r}
#| label: simulated-regression
# Fit and extract coefficients
sim_fit <- lm(fuel ~ temp, data = sim_data)

beta_0_sim <- coef(sim_fit)[1]
beta_1_sim <- coef(sim_fit)[2]
sigma_sq_sim <- sum(residuals(sim_fit)^2) / (8 - 2)
```

**Simulated model**

$$
\hat{\beta_0} = 14.61 \qquad \hat{\beta_1} = -0.108 \qquad \hat{\sigma^2} = 0.807
$$

---

<br>

```{r}
#| label: sim-data-fit
#|
# Plot original and simulated line
fuel |>
  ggplot(aes(x = temp, y = fuel)) +
  geom_point(alpha = 0.3) +
  geom_abline(
    intercept = beta_0,
    slope = beta_1,
    colour = ggokabeito::palette_okabe_ito(3)
  ) +
  geom_abline(
    intercept = beta_0_sim,
    slope = beta_1_sim,
    colour = ggokabeito::palette_okabe_ito(2),
    linetype = "dashed"
  ) +
  labs(x = "Temperature", y = "Fuel") +
  theme_st303()
```

## Let's do 500 simulations!

```{r}
#| label: big-simulation
#| echo: true
set.seed(54321) # Reprodicibility

# Simulate some data and fit model 500 times
sim_results <- purrr::map(1:500, function(i) {
  # Simulation -----------------------------------------------------------------
  x <- fuel$temp
  esp <- rnorm(n = n, mean = 0, sd = sqrt(sigma_sq))

  y <- beta_0 + (beta_1 * x) + esp

  sim_data <- tibble(temp = x, fuel = y)

  # Modelling ------------------------------------------------------------------
  sim_fit <- lm(fuel ~ temp, data = sim_data)

  beta_0_sim <- coef(sim_fit)[1]
  beta_1_sim <- coef(sim_fit)[2]
  sigma_sq_sim <- sum(residuals(sim_fit)^2) / (8 - 2)

  # Output ---------------------------------------------------------------------
  tibble(
    sample = i,
    beta_0 = beta_0_sim,
    beta_1 = beta_1_sim,
    sigma_sq = sigma_sq_sim,
    data = list(sim_data)
  )
}) |>
  bind_rows()
```

---

<br>

```{r}
#| label: fig-regression-lines
#| fig-cap: "500 simulated regression lines"
sim_lines <- sim_results |>
  select(sample, data) |>
  tidyr::unnest(data) |>
  split(~sample) |>
  purrr::imap(function(s, i) {
    fit <- lm(fuel ~ temp, data = s)

    broom::augment(fit) |> mutate(sample = i) |> relocate(sample)
  }) |>
  bind_rows()

# Plot the simulated lines
sim_lines |>
  ggplot(aes(x = temp, y = .fitted, group = sample)) +
  geom_line(linewidth = rel(0.5), alpha = 0.1) +
  labs(x = "Temperature", y = "Fuel") +
  theme_st303()
```

---

```{r}
#| label: fig-distribution
#| fig-cap: "Distribution of estimated parameters"
# Calculate mean ---------------------------------------------------------------
sim_plot <- sim_results |>
  select(-data) |>
  tidyr::pivot_longer(
    cols = !sample,
    names_to = "parameter",
    values_to = "value"
  ) |>
  group_by(parameter) |>
  mutate(mean_value = mean(value))

# Plotting ---------------------------------------------------------------------
sim_plot |>
  ggplot(aes(x = value, fill = parameter)) +
  geom_density(alpha = 0.5) +
  geom_vline(
    aes(xintercept = mean_value),
    linetype = "dashed",
    colour = "#999999"
  ) +
  ggokabeito::scale_fill_okabe_ito() +
  labs(x = NULL, y = "Density") +
  facet_wrap(
    ~parameter,
    scales = "free",
    labeller = labeller(
      parameter = function(x) {
        case_when(
          x == "beta_0" ~ "&beta;<sub>0</sub>",
          x == "beta_1" ~ "&beta;<sub>1</sub>",
          x == "sigma_sq" ~ "&sigma;<sup>2</sup>",
        )
      }
    )
  ) +
  guides(fill = "none") +
  theme_st303() +
  theme(strip.text = ggtext::element_markdown())
```

:::: {.fragment .fade-up}
::: {.callout-tip}
## These are called sampling distributions
:::
::::

## Sampling distribution of $\beta_1$ {style="font-size: 85%;"}

Under the assumptions of linear regression

$$
\boxed{\hat{\beta_1} \sim \mathcal{N}(\beta_1, \frac{\sigma^2}{S_{xx}})}
$$

where

$$
S_{xx} = \sum^n_{i=1}(x_i - \bar{x})^2
$$


::::: {.fragment}

[**Uncertainty in our estimate decreases when**]{.center}

:::: {.columns}

::: {.column}

[**More variation in $X$**]{.center}

$$
S_{xx} \uparrow
$$

:::

::: {.column}

[**Less unexplained variation**]{.center}

$$
\sigma^2 \downarrow
$$

:::

::::

:::::

## Sampling distribution of $\beta_0$

Under the assumptions of linear regression

$$
\boxed{\hat{\beta_0} \sim \mathcal{N}\Big(\beta_0, \sigma^2 \big( \frac{1}{n} + \frac{\bar{x^2}}{S_{xx}} \big) \Big)}
$$

[Here, the uncertainty in $\hat{\beta}_0$ depends on]{.fragment .fade-up}

:::{.incremental}
- $\sigma^2$: More noise means more uncertainty
- $n$: The larger the sample the less uncertain our estimate is
- $\bar{x}$ being close to 0
  - If $\bar{x}$ is far from zero, estimating the intercept becomes harder.
:::

## The sampling distribution of $\sigma^2$

$$
\sigma^2 = \frac{\sum \varepsilon_i^2}{n - 2}
$$

Thus, $\sigma^2$ would follow $\sigma^2 \sim \chi^2_{n - 2}$

```{r}
#| label: chi-dist
tibble(
  chi_6 = rchisq(n = 1000, df = 6),
  chi_8 = rchisq(n = 1000, df = 8),
  chi_10 = rchisq(n = 1000, df = 10),
) |>
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "df",
    values_to = "value"
  ) |>
  mutate(
    df = factor(df, levels = paste0("chi_", c(6, 8, 10)))
  ) |>
  ggplot(aes(x = value, fill = df)) +
  geom_density() +
  ggokabeito::scale_fill_okabe_ito() +
  labs(x = NULL, y = NULL) +
  facet_wrap(~df) +
  guides(fill = "none") +
  theme_st303()
```

:::{.notes}
- We subtract two because we are estimating two parameters $(\beta_0, \beta_1)$
:::

## Wrapping up {style="font-size: 85%;"}

- We can get estimates of our population parameters $\beta_0, \beta_1, \sigma^2$
- Errors are unobservable, while residuals are observable
- We make several assumptions about the errors
- Our parameters have known properties and distributions

```{r}
#| label: distributions
# Calculate mean ---------------------------------------------------------------
sim_plot <- sim_results |>
  select(-data) |>
  tidyr::pivot_longer(
    cols = !sample,
    names_to = "parameter",
    values_to = "value"
  ) |>
  group_by(parameter) |>
  mutate(mean_value = mean(value))

# Plotting ---------------------------------------------------------------------
sim_plot |>
  ggplot(aes(x = value, fill = parameter)) +
  geom_density(alpha = 0.5) +
  geom_vline(
    aes(xintercept = mean_value),
    linetype = "dashed",
    colour = "#999999"
  ) +
  ggokabeito::scale_fill_okabe_ito() +
  labs(x = NULL, y = "Density") +
  facet_wrap(
    ~parameter,
    scales = "free",
    labeller = labeller(
      parameter = function(x) {
        case_when(
          x == "beta_0" ~ "&beta;<sub>0</sub>",
          x == "beta_1" ~ "&beta;<sub>1</sub>",
          x == "sigma_sq" ~ "&sigma;<sup>2</sup>",
        )
      }
    )
  ) +
  guides(fill = "none") +
  theme_st303() +
  theme(strip.text = ggtext::element_markdown())
```

## Some useful R code for linear models

```{.r filename="useful-code.R"}
# Fit a simple linear regression model
fit <- lm(fuel ~ temp, data = fuel)

## View the model coefficients
coef(fit)

# View the intercept
coef(fit)[1]

# View the lope
coef(fit)[2]

# Get the full model summmary
summary(fit)

# View the residuals
residuals(fit)

# View the fitted values
fitted(fit)

# Make predictions from a fitted model
predict(fit, newdata = tibble(temp = 40))

# Or make predictions from a range of values
predict(fit, newdata = tibble(temp = c(30, 40, 50, 60))
```
