---
title: Inference for Simple Linear Regression
subtitle: ST303 - Linear Models
filters:
  - code-window
---

```{r}
#| label: setup

set.seed(4321)

# Output -----------------------------------------------------------------------
knitr::opts_chunk$set(fig.align = "center")

# Packages ---------------------------------------------------------------------
library(dplyr)
library(ggplot2)
library(tinytable)

# Theme ------------------------------------------------------------------------
theme_st303 <- function() {
  theme_bw() +
    theme(
      axis.title = element_text(size = rel(0.8), face = "bold"),
      legend.title.position = "top",
      legend.position = "bottom",
      legend.title = element_text(hjust = 0.5, size = rel(0.8), face = "bold"),
      panel.grid.minor = element_blank()
    )
}

# Data -------------------------------------------------------------------------
fuel <- readRDS(here::here("data/fuel.RDS"))

# Fit linear model -------------------------------------------------------------
fit <- lm(fuel ~ temp, data = fuel)

n <- nrow(fuel)
beta_0 <- coef(fit)[1]
beta_1 <- coef(fit)[2]
sigma_sq <- sum(residuals(fit)^2) / (n - 2)
```

## Recap of what we've done so far

So far, we have:

- We used **OLS** to estimate a regression line.
- We obtained estimates for $\beta_0$ and $\beta_1$
- We learned that these estimates would change if we collected a different sample.
- We learned about the **sampling distributions** of $\beta_0$ and $\beta_1$

<br>

::: {.callout-note}
## Today: How can we use these sampling distributions to make inferences
:::

## Three questions we might ask

**The slope**

What is the population value of $\beta_1$?

::: {.fragment .fade-up}
**The mean**

What is the average value of $Y$ when $X = x_0$?
:::


::: {.fragment .fade-up}
**An individual value**

What might one future value of $Y$ be when $X = x_0$?
:::

<br>

:::: {.fragment .fade-up}
::: {.callout-note icon="false"}
## These are different questions - so they involve different kinds of uncertainty
:::
::::

## Why do we need to do inference

Let's suppose our sample of data gave the following slope value

$$
\hat{\beta_1} = -0.128
$$


<br>

::: {.callout-important}
## Is this the true population slope?
:::


::: {.fragment .fade-up}
If we collected another sample, we would probably get a different estimate:

$$
\hat{\beta_1}^{(1)}, \quad \hat{\beta_1}^{(2)}, \quad \hat{\beta_1}^{(3)}, \quad \dots
$$

:::

<br>

:::: {.fragment .fade-up}
::: {.callout-tip}
## Inference is about quantifying this uncertainty.
:::
::::

## Standard error

One way of beginning to quantify that uncertainty is with the **standard error**

::: {.fragment .fade-up}

We know that the sampling variance of $\hat{\beta_1}$ is

$$
\text{Var}(\hat{\beta_1}) = \frac{\hat{\sigma}^2}{S_{xx}}
$$

:::


::: {.fragment .fade-up}

Therefore, the **standard error** is simply the square root of the variance:

$$
SE(\hat{\beta_1}) = \sqrt{\frac{\hat{\sigma}^2}{S_{xx}}} = \frac{\hat{\sigma}}{\sqrt{S_{xx}}}
$$

:::

---

<br>

$$
SE(\hat{\beta_1}) = \frac{\hat{\sigma}}{\sqrt{S_{xx}}}
$$

[**So uncertainty decreases when**]{.center}

:::: {.columns}

::: {.column}

[**Less residual variation**]{.center}

$$
\hat{\sigma} \downarrow
$$

:::

::: {.column}

[**More information about $X$**]{.center}

$$
S_{xx} \uparrow
$$

:::
::::

:::: {.fragment .fade-up}
::: {.callout-tip}
## The more information we have, the more precise our estimates will be
:::
::::

## Quick question

If we **increase the sample size** while keeping everything else similar

What would you expect to happen to

$$
SE(\hat{\beta_1})
$$

:::: {.columns}
::: {.column width="33%"}
[**Increase**]{.center}
:::

::: {.column width="33%"}
[**Stay the same**]{.center}
:::

::: {.column width="33%"}
[**Decrease**]{.center .fragment .highlight-green}
:::
::::


:::: {.fragment .fade-up}

$$
\boxed{SE(\hat{\beta_1}) \downarrow}
$$

::: {.callout-tip appearance="minimal"}
## More observations generally give us a more precise estimate

More data tends to increase $S_{xx}$ while $\hat{\sigma}$ stays roughly stable
:::
::::

## Uncertainty → inference

So far we have

- An estimate of our slope: $\hat{\beta_1}$
- An estimate of it's uncertainty: $\text{SE}(\hat{\beta_1})$

Now we can ask

::: {.callout-note}
## What values of $\beta_1$ are compatible with our data?
:::

<br>

:::: {.fragment .fade-up}
::: {.callout-important}
## This gives us a confidence interval
:::
::::

## Recall our assumptions

Before we build intervals, recall the model assumptions:

1. **Linearity**: the mean of $Y$ is a linear function of $X$
2. **Independence**: observations are independent
3. **Constant variance**: $\text{Var}(\varepsilon_i) = \sigma^2$ for all $i$
4. **Normality**: $\varepsilon_i \sim \mathcal{N}(0, \sigma^2)$

<br>

::: {.callout-note icon="false"}
## SEs, $t$-tests, CIs, PIs all rely on these being reasonable
:::

## Confidence intervals

<br>

$$
\boxed{\text{estimate } \pm \text{ critical value } \times \text{ standard error}}
$$

<br>

For our slope this is

$$
\boxed{\hat{\beta_1} \pm t_{n-2, 1 - \alpha/2} \; \text{SE}(\hat{\beta_1})}
$$

where

$$
\alpha = 0.05
$$

## Why use the $t$-distribution

If we knew the true value of $\sigma$

$$
\frac{\hat{\beta_1} - \beta_1}{\sigma / \sqrt{S_{xx}}} \sim \mathcal{N}(0, 1)
$$

::: {.fragment .fade-up}

But we only have $\hat{\sigma}$, and estimating it introduces
**additional uncertainty**.

Hence we use a $t_{n-2}$ distribution:

$$
\frac{\hat{\beta_1} - \beta_1}{\hat{\sigma} / \sqrt{S_{xx}}} \sim t_{n - 2}
$$

Compared with the standard normal distribution, the $t$-distribution has
**heavier tails**
:::

## Let's look at the fuel data example {style="font-size: 75%;"}

### Confidence interval for the slope

From previous lectures we know that

$$
\hat{\beta_1} = -0.128 \qquad \hat{\sigma^2} = 0.428 \qquad S_{xx} = 1404.355
$$

Therefore the standard error of $\hat{\beta}_1$ is:

$$
SE(\hat{\beta_1}) = \frac{\hat{\sigma}}{\sqrt{S_{xx}}} = \sqrt{\frac{0.428}{1404.355}} \approx 0.018
$$

And from *Student’s t-Table* we know that

$$
t_{(n-2), (1 - \alpha/2)} = t_{6, 0.975} = 2.447 \qquad \text{where } n = 8 \; \& \; \alpha = 0.05
$$


:::: {.fragment .fade-up}

$$
\text{CI}(\hat{\beta_1}) = -0.128 \pm (2.447 \times 0.018) = (-0.172, -0.084)
$$


::: {.callout-tip}
## We are 95% confident that the population slope is between $-0.172$ and $-0.084$.
:::
::::

## Quick little side note

### Observed t-values versus critical t-values

```{r}
#| label: model-summary
#| echo: true
#| collapse: true
summary(fit)
```

## Performing inference on $\beta_1$ {style="font-size: 75%;"}

In some cases we may wish to perform a hypothesis test

$$
H_0: \beta_1 = 0 \qquad H_a: \beta_1 \ne 0
$$

Under the null hypothesis $(H_0)$

$$
t = \frac{\hat{\beta_1} - 0}{SE(\hat{\beta_1})}
$$


:::: {.fragment}

From our data

$$
t = \frac{-0.128}{0.018} = -7.11 \qquad \text{df } = 6
$$

Hence we get $p < 0.001$

::: {.callout-important}
## There is strong evidence that the population slope is not zero.
:::
::::

## Performing inference on $\beta_1$

### Using the confidence intervals

For a two-sided test at $\alpha = 0.05$ with $H_0: \beta_1 = 0$

We **reject $H_0$** when the 95% confidence interval **does not** cross 0

:::: {.fragment .fade-up}
Our confidence interval was

$$
(-0.172, -0.084) \notin 0
$$

::: {.callout-important}
## Providing further evidence for rejecting $H_0$
:::
::::

## Performing inference on $\beta_0$

We can apply the logic for $\beta_0$

$$
SE(\hat{\beta}_0) = \hat{\sigma} \sqrt{\frac{1}{n} + \frac{\bar{x}^2}{S_{xx}}}
$$

$$
\hat{\beta}_0 \pm t_{(n-2),\, (1-\alpha/2)} \; SE(\hat{\beta}_0)
$$


::: {.fragment .fade-up}
Therefore, for our fuel data the 95% CI for $\beta_0$ is:

$$
95\% (CI) = (13.88, 17.80)
$$

:::


:::: {.fragment}
::: {.callout-note}
## We are 95% confident that the population intercept is between $13.88$ and $17.80$
:::
::::

## Making predictions from linear models

### Mean versus individual predictions

Let's say we want to make a prediction from our model

$$
\hat{y} = 15.84 - 0.128(50) = 9.44
$$

At $50^\circ$F, predicted fuel usage is 9.44.

[**But this value can be interpreted in two ways***]{.fragment .fade-up .center}


:::: {.fragment .fade-up}
::: {.callout-warning appearance="minimal"}
## 9.44 is the [average fuel use across all weeks]{.alert} where temperature is $50^\circ$F
:::
::::

:::: {.fragment .fade-up}
::: {.callout-warning appearance="minimal"}
## 9.44 is the fuel usage [for one particular week]{.alert} where temperature is $50^\circ$F
:::
::::

:::: {.fragment}
::: {.callout-tip appearance="minimal"}
## The former has a [confidence interval]{.alert} and the latter has a [prediction interval]{.alert}
:::
::::

## Confidence versus prediction intervals

::: {.callout-note appearance="minimal" title="Confidence versus prediction intervals"}
- A confidence interval is concerned with how uncertain are we <br> about the
  **average response** at this value of $X$?
- A prediction interval is concerned with how uncertain are we <br> about one
  **particular future response** at this value of $X$?
:::

::: {.fragment .fade-up}
::: {.callout-tip appearance="minimal"}
## A prediction interval has [two]{.alert} sources of uncertainty

- Uncertainty about where the regression line is
- Natural variation of individual observations around the line.
:::
:::

::: {.fragment .fade-up}

- The confidence interval is **about where the line** is
- The prediction interval is **about where an individual observation might land**

:::

## Confidence interval for the mean response {style="font-size: 95%;"}

::: {.callout-note icon="false"}
## What is the average fuel use for weeks where temperature is $50^\circ$F?
:::

We know that $\hat{y_0} = 9.44$ and the standard error for this estimate can be
represented as:

$$
SE({\hat{y_0}}) = \hat{\sigma} \sqrt{\frac{1}{n} + \frac{(x_0 -\bar{x})^2}{S_{xx}}}
$$


::: {.fragment .fade-up}

$$
\begin{align}
\text{Hence } 95\% \text{ CI} &= 9.44 \; \pm \; 2.447 \times \color{blue}{0.254} \\[4pt]
&= (8.82, 10.06)
\end{align}
$$

:::

::: {.fragment .fade-up}
::: {.callout-tip appearance="minimal"}
## Our interpretation

We are 95% confident that the mean fuel use at $50^\circ F$ is between approximately 8.82 and 10.06 units.
:::
:::

## Prediction intervals

::: {.callout-note icon="false"}
## What might the fuel use be for one particular week when temperature is $50^\circ$F?
:::

We know that $\hat{y_0} = 9.44$, but this time we have to account for additional
uncertainty around the regression line

$$
SE_{\text{prediction}} = \hat{\sigma} \sqrt{1 + \frac{1}{n} + \frac{(x_0 -\bar{x})^2}{S_{xx}}}
$$


::: {.fragment .fade-up}
$$
\begin{align}
\text{Hence } 95\% \text{ CI} &= 9.44 \; \pm \; 2.447 \times \color{blue}{0.702} \\[4pt]
&= (7.72, 11.16)
\end{align}
$$

:::



:::: {.fragment .fade-up}
::: {.callout-important appearance="minimal"}
## Notice how to PI is much wider than the CI

It must cover both our uncertainty about the line *and* the natural spread
of individual observations
:::
::::

---

<br>

```{r}
# Generate a new data
new_data <- tibble(temp = seq(28, 63, by = 0.1))

# Doing confidence intervals
preds_1 <- predict(fit, newdata = new_data, interval = "confidence") |>
  as_tibble() |>
  mutate(interval = "Confidence interval") |>
  bind_cols(new_data)

# Doing prediction intervals
preds_2 <- predict(fit, newdata = new_data, interval = "prediction") |>
  as_tibble() |>
  mutate(interval = "Prediction interval") |>
  bind_cols(new_data)

# Joining them both together
preds_full <- bind_rows(preds_1, preds_2)

# Plotting
preds_full |>
  ggplot(aes(x = temp, y = fit)) +
  geom_line(linewidth = rel(1)) +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr, fill = interval, colour = interval),
    alpha = 0.2
  ) +
  ggokabeito::scale_fill_okabe_ito() +
  ggokabeito::scale_colour_okabe_ito() +
  labs(
    x = "Temperature",
    y = "Predicted fuel use",
    fill = NULL,
    colour = NULL
  ) +
  theme_st303()
```

## Doing predictions in R

```{r}
#| label: predictions
#| echo: true

# Doing predictions with a confidence interval
predict(fit, newdata = data.frame(temp = 50), interval = "confidence")

# Doing predictions with a prediction interval
predict(fit, newdata = data.frame(temp = 50), interval = "prediction")
```

## Summary: one formula, four intervals {style="font-size: 90%;"}

Every interval we've seen has the form $\text{estimate} \pm t^* \times SE$:

<br>
<br>

| Quantity | Estimate | Standard error |
|---|---|---|
| $\beta_1$ | $\hat{\beta}_1$ | $\hat{\sigma} / \sqrt{S_{xx}}$ |
| $\beta_0$ | $\hat{\beta}_0$ | $\hat{\sigma}\sqrt{\tfrac{1}{n} + \tfrac{\bar{x}^2}{S_{xx}}}$ |
| Mean response at $x_0$ | $\hat{y}_0$ | $\hat{\sigma}\sqrt{\tfrac{1}{n} + \tfrac{(x_0-\bar{x})^2}{S_{xx}}}$ |
| Individual response at $x_0$ | $\hat{y}_0$ | $\hat{\sigma}\sqrt{1 + \tfrac{1}{n} + \tfrac{(x_0-\bar{x})^2}{S_{xx}}}$ |

<br>

::: {.callout-tip}
## The only difference between a CI and a PI is the extra "$1 +$" under the square root.
:::
