---
title: Linear models and ordinary least squares
page-layout: full
---

```{r}
#| label: setup
#| include: false

purrr::map(list.files(here::here("R"), full.names = TRUE), source)
```

## Learning objectives

- Understand where residuals come from
- Understand the intercept and slope
- Learn how to derive the formula for the intercept and slope

## Readings

### Suggested readings

- {{< fa book >}} [Linear models in statistics (Chapter 6)](https://utstat.utoronto.ca/~brunner/books/LinearModelsInStatistics.pdf)
- {{< fa link >}} [Ordinary Least Squares Regression - Explained Visually](https://setosa.io/ev/ordinary-least-squares-regression/)
- {{< fa link >}} [Doing linear regression by hand](https://c-monaghan.github.io/posts/2025/03/12/Linear-Regression/)

## Slides

### Lecture 1

::: {.callout-caution}
## Slides will be available on Friday 25th September
:::

```{r}
#| label: week-1a
#| echo: false
#| eval: false
#| results: asis

add_slides(slide_id = "/slides/output/slide_1.html")
```

### Lecture 2

::: {.callout-caution}
## Slides will be available on Friday 25th September
:::

```{r}
#| label: week-1b
#| echo: false
#| eval: false
#| results: asis

add_slides(slide_id = "/slides/output/slide_2.html")
```
