---
title: Welcome to ST303
page-layout: full
---

```{r}
#| label: setup
#| include: false

# Functions --------------------------------------------------------------------
purrr::map(list.files(here::here("R"), full.names = TRUE), source)

# Class timetable --------------------------------------------------------------
timetable <- tibble::tribble(
  ~Time                 , ~Location                , ~`Class Details`           ,
  "Friday (12pm - 1pm)" , "Arts Building (Room F)" , "Standard format (slides)" ,
  "Friday (4pm - 5pm)"  , "Physics Hall"           , "Standard format (slides)"
)
```

## Module overview

Welcome to third year and welcome to ST303!

We will be covering lots of things related to linear models. This will range
from the simple and multiple linear regression, ordinary and weighted least
squares, inference, and diagnostics (among other things).

### Learning outcomes

On successful completion of the module, you should be able to

- Fit and interpret multiple regression models.
- Understand the theory of multiple regression.
- Assess model assumptions.
- Use model diagnostics.

### Assessment

This module is broken down as 25% continuous assessment and 75% official exam.

- 4 assignments in the form of a **Moodle quiz** (worth 10%)
- 1 mid-term exam (worth 15%)
- 1 university exam (worth 75%)

::: {.callout-note icon="false"}
## The dates for each exam are still being finalised
:::

::: {.callout-important title="Details for Erasmus students"}
If you are an Erasmus student here for **Semester 1 only**
(September - December), you will not be here for the final exam. With that in
mind, your assignments and mid-term will be worth more marks.

If you are here for the full academic year (semester 1 **AND** semester 2), then
your grades will follow the same breakdown as listed above.
:::

### Timetable

```{r}
#| label: timetable
#| echo: false
#| results: asis
timetable |>
  kableExtra::kbl(
    escape = FALSE,
    align = "llc",
    table.attr = 'class="schedule-table"'
  ) |>
  kableExtra::kable_styling() |>
  kableExtra::column_spec(
    1,
    width = "40%",
    extra_css = "padding-right: 20px;"
  ) |>
  kableExtra::column_spec(2:3, width = "30%")
```

### Tutorials

- Tutorials will run weekly starting from the **week of the 5th October (week 3)**
- These tutorials will be in the form of help desks where you get assistance
  with the assignments and consolidate and apply what you learned in class.

## Slides

```{r}
#| label: add-slides
#| echo: false
#| results: asis

add_slides(slide_id = "/slides/output/slide_0.html")
```
