---
title: Practice Questions Sheet 1
subtitle: ST303 - Linear Models
author: Cormac Monaghan
description: |
  This sheet provides practice with simple linear regression. You will work
  with the regression model algebraically, fit and interpret a model in `R`,
  and reproduce key regression calculations by hand. These practice questions
  should be attempted in your own time. These questions are **not graded** and **should not be handed in**

  You should attend the joint ST302/ST303 lab to get help with the questions
  and have your answers checked.
format:
  st303_assignment-html: default
  st303_assignment-typst: default
---

```{r}
#| label: setup
#| include: false

# Computations -----------------------------------------------------------------
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)

# Packages ---------------------------------------------------------------------
library(dplyr)
library(tinytable)

# Example data -----------------------------------------------------------------
concrete <- tibble(
  `7_day` = c(2300, 2430, 2890, 3120, 3380, 3390),
  `28_day` = c(4370, 4640, 4620, 4900, 5020, 5220)
)
```

\newpage

# Simple linear regression

Consider the following simple linear regression model
$$
y_i = \beta_0 + \beta_1x_i + \epsilon_i, \qquad i = 1,\ldots,n.
$$

The above model may be alternatively written as

$$
y_i = \alpha_0 + \alpha_1(x_i-\bar{x}) + \epsilon_i, \qquad i = 1,\ldots,n.
$$

Answer the following:

1. Using the method of least squares, derive expressions for the estimators
   $\hat{\alpha}_0$ and $\hat{\alpha}_1$
2. Interpret the parameters $\alpha_0$ and $\alpha_1$.
3. Explain why mean-centering the predictor variable can be useful.

\newpage

# Simple linear regression in `R`

As concrete cures, it gains strength overtime. The following data
(see @tbl-concrete-data) represent the 7-day and 28-day strength, in pounds
per square inch (psi), of a certain type of concrete.

```{r}
#| label: tbl-concrete-data
#| tbl-cap: "Concrete data"
concrete |>
  rename(`7 Day` = `7_day`, `28 Day` = `28_day`) |>
  tt() |>
  style_tt(i = 0, bold = TRUE) |>
  style_tt(j = 1:2, align = "c")
```

You can read the data from the RStudio server using:

```{r}
#| label: load-data
#| eval: false
#| echo: true

# Using the `readr` and the `here` package.
# TIP: you can press tab when writing the folder path to have all the options
# folder / file options appear to you

concrete <- readr::read_csv(here::here("Sharedfiles/ST303/data/Concrete.csv"))
```

Answer the following set of questions using `R`

1. Create a scatterplot showing the relationship between 7-day strength
  on the $x$-axis and 28-day strength on the $y$-axis.
2. Does it seem appropriate to assume a linear relationship between the
  two variables (explain why you think so)?
3. Specify a simple linear regression model for these data algebraically.
4. Fit this model in `R` and report what the **intercept**, **slope**, and
  **standard error** for each term are.
5. In the context of this data, what do these values mean?
6. Update the scatter plot you made earlier to include a fitted line.

\newpage

# Simple linear regression by hand

This time, instead of using `R`, we will be calculating our linear regression
parameters by hand. Using the concrete data we mentioned in @tbl-concrete-data,
calculate the following:

:::{.content-visible when-format="typst"}

```{=typst}
#v(0.25cm)
```

:::

1. $\bar{x}$, $\sum_{i=1}^n x_i^2$, and $S_{xx}$.
2. $\bar{y}$, $\sum_{i=1}^n y_i^2$, and $S_{yy}$.
3. $\sum_{i=1}^n x_i y_i$ and $S_{xy}$.

:::{.content-visible when-format="typst"}

```{=typst}
#v(0.25cm)

#box(stroke: rgb("#667477"), inset: 1em)[#text(fill: rgb("#667477"))[
  You may use `R` to verify your calculations.

  However, in an exam setting you may be asked to do calculations by
  hand (_hint, hint, wink, wink!!_).

  *As such, please show all of your workings*.
]]

#v(0.25cm)
```

:::

:::: {.content-visible when-format="html"}
::: {.callout-tip appearance="simple"}
## You may use `R` to verify your calculations.

However, in an exam setting you may be asked to do calculations by hand
(*hint, hint, wink, wink!!*).

**As such, please show all of your workings**.
:::
::::

Use your calculated quantities to:

4. Calculate the least-squares slope estimate and the least-squares intercept
  estimate
5. In addition, calculate the estimated error variance.
